Exploring Local Explanations of Nonlinear Models Using Animated Linear Projections

### 2022-04-25 _paper.rmd
few grammerly edits

<!-- Paper structure -->
The remainder of this paper is organized as follows. Section \@ref(sec:explanations) covers the background of the local explanation and the traditional visuals produced. Section \@ref(sec:tour) explains the animations of continuous linear projections. Section \@ref(sec:cheemviewer) discusses the visual layout in the graphical user interface, and how it facilitates analysis, data preprocessing, and package infrastructure. Then Section \@ref(sec:casestudies) illustrates the application of supervised learning with categorical and quantitative output. Section \@ref(sec:cheemdiscussion) concludes with the insights gained and possible directions to explore in the future.

@@

<!-- Purpose -->
The global view provides context for all instances and facilitates the exploration of the separability of the data- and attribution-spaces. These spaces both have dimensionality $n \times p$, where $n$ is the number of instances and $p$ is the number of featuress. The attribution space corresponds to the local explanations for each instance; feature importance in the vicinity of the instance.

@@

Follow the examples provided with the package to compute the local explaination (using `?cheem_ls`). The application expects the output returned by `cheem_ls()`, saved to an `rds` file with `saveRDS()` to be uploaded.

@@

<!-- Dark misclassified as milk, rational -->
It could be interesting to examine the nutritional properties of any dark chocolates that have been misclassified as milk. A reason to do this is that a dark chocolate nutritionally more like milk should not be considered a healthy alternative. It is interesting to explore which nutritional features contribute most to misclassification.

@@

<!-- Attention to Density, end of first case -->
From the density plot, which is the attribution projection corresponding to the PI, both instances are more like dark chocolates. Varying the contribution of Sugars and altogether removing it from the projection is where the difference becomes apparent. When a frame with contribution primarily of Fiber is examined instance 22 looks more like a milk chocolate.

@@

Ames housing data 2018 was subset to North Ames (the neighborhood with the most house sales). The remaining are 338 house sales. A random forest model was fit predicting the sale price [USD] from the property features: Lot Area (`LtA`), Overall Quality (`Qlt`), Year the house was Built (`YrB`), Living Area (`LvA`), number of Bathrooms (`Bth`), number of Bedrooms (`Bdr`), the total number of Rooms (`Rms`), Year the Garage was Built (`GYB`), and Garage Area (`GrA`). Using  interactions with the global view, a house with an extreme negative residual and an accurate instance with a similar prediction is selected.

(ref:caseames-cap) The largest neighborhood of Ames house price data is fit with a random forest predicting the sales price [USD]. Exploring an instance with a extreme residual as the PI and an accurate prediction for a similar price is used for comparison. The local explanation indicates a sizable attribution to Lot Area (`LtA`), while the CI has minimal attribution to this feature. The PI has a higher predicted value than the CI in the attribution projection. Reducing the contribution of Lot Area brings these two prices in line. This suggests that if the model did not value Lot Area so highly for this observation, then the observed sales price would be quite similar. That is, the large residual is due to a lack of factoring in the Lot Area for the prediction of PI's sales price. (A video showing the animation is at \url{https://vimeo.com/666431134}.)
```{r caseames, echo=F, out.width="90%", fig.cap = "(ref:caseames-cap)"}
knitr::include_graphics("./figures/case_ames2018.png")
```

>>>>

<!-- Paper structure -->
The remainder of this paper is organized as follows. Section \@ref(sec:explanations) covers the background of the local explanation and the traditional visuals produced. Section \@ref(sec:tour) explains the animations of continuous linear projections. Section \@ref(sec:cheemviewer) discusses the visual layout in the graphical user interface and how it facilitates analysis, data preprocessing, and package infrastructure. Then Section \@ref(sec:casestudies) illustrates the application of supervised learning with categorical and quantitative output. Section \@ref(sec:cheemdiscussion) concludes with the insights gained and possible directions to explore in the future.

@@

<!-- Purpose -->
The global view provides context for all instances and facilitates the exploration of the separability of the data- and attribution-spaces. These spaces both have dimensionality $n \times p$, where $n$ is the number of instances and $p$ is the number of features. The attribution space corresponds to the local explanations for each instance; feature importance in the vicinity of the instance.

@@

Follow the examples provided with the package to compute the local explanation (using `?cheem_ls`). The application expects the output returned by `cheem_ls()`, saved to an `rds` file with `saveRDS()` to be uploaded.

@@

<!-- Dark misclassified as milk, rational -->
It could be interesting to examine the nutritional properties of any dark chocolates that have been misclassified as milk. A reason to do this is that a dark chocolate, nutritionally more like milk should not be considered a healthy alternative. It is interesting to explore which nutritional features contribute most to misclassification.

@@

<!-- Attention to Density, end of first case -->
From the density plot, which is the attribution projection corresponding to the PI, both instances are more like dark chocolates. Varying the contribution of Sugars and altogether removing it from the projection is where the difference becomes apparent. When a frame with contribution primarily from Fiber is examined instance 22 looks more like a milk chocolate.

@@

Ames housing data 2018 was subset to North Ames (the neighborhood with the most house sales). The remaining are 338 house sales. A random forest model was fit, predicting the sale price [USD] from the property features: Lot Area (`LtA`), Overall Quality (`Qlt`), Year the house was Built (`YrB`), Living Area (`LvA`), number of Bathrooms (`Bth`), number of Bedrooms (`Bdr`), the total number of Rooms (`Rms`), Year the Garage was Built (`GYB`), and Garage Area (`GrA`). Using  interactions with the global view, a house with an extreme negative residual and an accurate instance with a similar prediction is selected.

(ref:caseames-cap) The largest neighborhood of Ames house price data is fit with a random forest predicting the sales price [USD]. Exploring an instance with an extreme residual as the PI and an accurate prediction for a similar price is used for comparison. The local explanation indicates a sizable attribution to Lot Area (`LtA`), while the CI has minimal attribution to this feature. The PI has a higher predicted value than the CI in the attribution projection. Reducing the contribution of Lot Area brings these two prices in line. This suggests that if the model did not value Lot Area so highly for this observation, then the observed sales price would be quite similar. That is, the large residual is due to a lack of factoring in the Lot Area for the prediction of PI's sales price. (A video showing the animation is at \url{https://vimeo.com/666431134}.)
```{r caseames, echo=F, out.width="90%", fig.cap = "(ref:caseames-cap)"}
knitr::include_graphics("./figures/case_ames2018.png")
```

### 2022-04-25 _paper.tex
few grammerly edits

The remainder of this paper is organized as follows. Section \ref{sec:explanations} covers the background of the local explanation and the traditional visuals produced. Section \ref{sec:tour} explains the animations of continuous linear projections. Section \ref{sec:cheemviewer} discusses the visual layout in the graphical user interface, and how it facilitates analysis, data preprocessing, and package infrastructure. Then Section \ref{sec:casestudies} illustrates the application of supervised learning with categorical and quantitative output. Section \ref{sec:cheemdiscussion} concludes with the insights gained and possible directions to explore in the future.

@@

The global view provides context for all instances and facilitates the exploration of the separability of the data- and attribution-spaces. These spaces both have dimensionality \(n \times p\), where \(n\) is the number of instances and \(p\) is the number of featuress. The attribution space corresponds to the local explanations for each instance; feature importance in the vicinity of the instance.

@@

Follow the examples provided with the package to compute the local explaination (using \texttt{?cheem\_ls}). The application expects the output returned by \texttt{cheem\_ls()}, saved to an \texttt{rds} file with \texttt{saveRDS()} to be uploaded.

@@

It could be interesting to examine the nutritional properties of any dark chocolates that have been misclassified as milk. A reason to do this is that a dark chocolate nutritionally more like milk should not be considered a healthy alternative. It is interesting to explore which nutritional features contribute most to misclassification.

@@

From the density plot, which is the attribution projection corresponding to the PI, both instances are more like dark chocolates. Varying the contribution of Sugars and altogether removing it from the projection is where the difference becomes apparent. When a frame with contribution primarily of Fiber is examined instance 22 looks more like a milk chocolate.

@@

Ames housing data 2018 was subset to North Ames (the neighborhood with the most house sales). The remaining are 338 house sales. A random forest model was fit predicting the sale price {[}USD{]} from the property features: Lot Area (\texttt{LtA}), Overall Quality (\texttt{Qlt}), Year the house was Built (\texttt{YrB}), Living Area (\texttt{LvA}), number of Bathrooms (\texttt{Bth}), number of Bedrooms (\texttt{Bdr}), the total number of Rooms (\texttt{Rms}), Year the Garage was Built (\texttt{GYB}), and Garage Area (\texttt{GrA}). Using interactions with the global view, a house with an extreme negative residual and an accurate instance with a similar prediction is selected.

@@

\caption{The largest neighborhood of Ames house price data is fit with a random forest predicting the sales price {[}USD{]}. Exploring an instance with a extreme residual as the PI and an accurate prediction for a similar price is used for comparison. The local explanation indicates a sizable attribution to Lot Area (\texttt{LtA}), while the CI has minimal attribution to this feature. The PI has a higher predicted value than the CI in the attribution projection. Reducing the contribution of Lot Area brings these two prices in line. This suggests that if the model did not value Lot Area so highly for this observation, then the observed sales price would be quite similar. That is, the large residual is due to a lack of factoring in the Lot Area for the prediction of PI's sales price. (A video showing the animation is at \url{https://vimeo.com/666431134}.)}\label{fig:caseames}
\end{figure}

>>>>

The remainder of this paper is organized as follows. Section \ref{sec:explanations} covers the background of the local explanation and the traditional visuals produced. Section \ref{sec:tour} explains the animations of continuous linear projections. Section \ref{sec:cheemviewer} discusses the visual layout in the graphical user interface and how it facilitates analysis, data preprocessing, and package infrastructure. Then Section \ref{sec:casestudies} illustrates the application of supervised learning with categorical and quantitative output. Section \ref{sec:cheemdiscussion} concludes with the insights gained and possible directions to explore in the future.

@@

The global view provides context for all instances and facilitates the exploration of the separability of the data- and attribution-spaces. These spaces both have dimensionality \(n \times p\), where \(n\) is the number of instances and \(p\) is the number of features. The attribution space corresponds to the local explanations for each instance; feature importance in the vicinity of the instance.

@@

Follow the examples provided with the package to compute the local explanation (using \texttt{?cheem\_ls}). The application expects the output returned by \texttt{cheem\_ls()}, saved to an \texttt{rds} file with \texttt{saveRDS()} to be uploaded.

@@

It could be interesting to examine the nutritional properties of any dark chocolates that have been misclassified as milk. A reason to do this is that a dark chocolate, nutritionally more like milk should not be considered a healthy alternative. It is interesting to explore which nutritional features contribute most to misclassification.

@@

From the density plot, which is the attribution projection corresponding to the PI, both instances are more like dark chocolates. Varying the contribution of Sugars and altogether removing it from the projection is where the difference becomes apparent. When a frame with contribution primarily from Fiber is examined instance 22 looks more like a milk chocolate.

@@

Ames housing data 2018 was subset to North Ames (the neighborhood with the most house sales). The remaining are 338 house sales. A random forest model was fit, predicting the sale price {[}USD{]} from the property features: Lot Area (\texttt{LtA}), Overall Quality (\texttt{Qlt}), Year the house was Built (\texttt{YrB}), Living Area (\texttt{LvA}), number of Bathrooms (\texttt{Bth}), number of Bedrooms (\texttt{Bdr}), the total number of Rooms (\texttt{Rms}), Year the Garage was Built (\texttt{GYB}), and Garage Area (\texttt{GrA}). Using interactions with the global view, a house with an extreme negative residual and an accurate instance with a similar prediction is selected.

@@

\caption{The largest neighborhood of Ames house price data is fit with a random forest predicting the sales price {[}USD{]}. Exploring an instance with an extreme residual as the PI and an accurate prediction for a similar price is used for comparison. The local explanation indicates a sizable attribution to Lot Area (\texttt{LtA}), while the CI has minimal attribution to this feature. The PI has a higher predicted value than the CI in the attribution projection. Reducing the contribution of Lot Area brings these two prices in line. This suggests that if the model did not value Lot Area so highly for this observation, then the observed sales price would be quite similar. That is, the large residual is due to a lack of factoring in the Lot Area for the prediction of PI's sales price. (A video showing the animation is at \url{https://vimeo.com/666431134}.)}\label{fig:caseames}
\end{figure}
